The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-03-18 script/en/chapters/primes.tex
CT Book ed 12: Cover as png instead of pdf (smaller size); added chapter 1.1; refreshed Mersenne records; corrected some minor typos.

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, Jan. 2010, Aug. 2013)

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 20+ largest known primes (as of February 2013)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{prime number!records}

The following table contains the biggest, currently known primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet at
     \url{http://primes.utm.edu/largest.html}
and at
     \url{http://primes.utm.edu/mersenne/index.html}.
}:

@@

\begin{table} % don't use [h] (error if table has >24 lines)
\begin{center}
\begin{tabular}{|c|cccc|}
\hline    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline
	1  & $2^{57,885,161}-1$ & 17,425,170 & 2013 & Mersenne, 48th known \\
	2  & $2^{43,112,609}-1$ & 12,978,189 & 2008 & Mersenne, 47th known \\
	3  & $2^{42,643,801}-1$ & 12,837,064 & 2009 & Mersenne, 46th known \\
	4  & $2^{37,156,667}-1$ & 11,185,272 & 2008 & Mersenne, 45th known \\
	5  & $2^{32,582,657}-1$ &  9,808,358 & 2006 & Mersenne, 44th known \\
	6  & $2^{30,402,457}-1$ &  9,152,052 & 2005 & Mersenne, 43rd known \\
	7  & $2^{25,964,951}-1$ &  7,816,230 & 2005 & Mersenne, M-42 \\
	8  & $2^{24,036,583}-1$ &  7,235,733 & 2004 & Mersenne, M-41 \\
	9  & $2^{20,996,011}-1$ &  6,320,430 & 2003 & Mersenne, M-40 \\
	10 & $2^{13,466,917}-1$ &  4,053,946 & 2001 & Mersenne, M-39 \\
	11 & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\
	12 & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\
	13 & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\
	14  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
	15  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
	16  & $4,847 \cdot 2^{3,321,063}+1$ & 999,744 & 2005 & Generalized Mersenne \\
	17  & $3 \cdot 2^{3,136,255}-1$ & 944,108 & 2007 & Generalized Mersenne \\
	18  & $2^{ 3,021,377}-1$ &   909,526 & 1998 & Mersenne, M-37 \\
	19  & $2^{ 2,976,221}-1$ &   895,932 & 1997 & Mersenne, M-36 \\
	20  & $222,361 \cdot 2^{2,854,840}+1$ & 859,398 & 2006 & Generalized Mersenne \\
	21 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\
	22  & $1,361,244^{131,072}+1$ &   803,988 & 2004 & Generalized Fermat \\
	23  & $1,176,694^{131,072}+1$ &   795,695 & 2003 & Generalized Fermat \\
	24  & $342,673 \cdot 2^{2,639,439}-1$ & 794,556 & 2007 & Generalized Mersenne \\

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of February 2013)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\addtocounter{footnote}{-1}
\footnotetext{\index{number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\url{http://www.seventeenorbust.com})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its task sometime.

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
Landon Curt Noll\index{Noll, Landon Curt} lists all known Mersenne primes
including its date of discovery and its value as number and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html} \\
Also see:
      \url{http://www.utm.edu/}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 42,643,801, ~ 43,112,609.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$48$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{prime number!Mersenne}\index{Mersenne!prime number}.

@@

\vskip +10 pt
For the first 42  % be_2005_UPDATEN_if-new-mersenne-prime-appears  % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 48th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely.\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/status.htm}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{prime number!test}. }

% be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Right now (February 2013) all prime exponents smaller than $ 26,083,693 $ have been
tested and double-checked\footnote{%
See home page of the GIMPS project:
      \url{http://www.mersenne.org/report_milestones}.
                                  }:
So we can
be certain, that this is really the 41th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime). Here are some examples in more detail:

@@

Right now the GIMPS project has discovered 14
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
largest Mersenne primes so far, including the largest known prime number at all.

@@

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\begin{table}[ht]
\begin{center}
\begin{tabular}{|cccc|}
\hline
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline

	$2^{57,885,161}-1$ & 17.425.170 & January 25, 2013 & Curtis Cooper  \\
	$2^{43,112,609}-1$ & 12,978,189 & August 23, 2008 & Edson Smith  \\
	$2^{42,643,801}-1$ & 12,837,064 & April 12, 2009 & Odd Magnar Strindmo \\
	$2^{37,156,667}-1$ & 11,185,272 & September 6, 2008 & Hans-Michael Elvenich \\

@@

\hline
\end{tabular}
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\caption{The largest primes found by the GIMPS project (as of February 2013)}
\label{Primes_GIMPS-Primes_table-reference}
\end{center}
\end{table}

@@

Formulae for the prime number theorem only apply when n tends to infinity,
The formula of Gausee can be replaced by more precise formulae.
For $x \geq 67$:
$$ ln(x) - 1,5 < x / PI(x) < ln(x) - 0,5 $$
Given that we know $PI(x)  =  x / \ln x$ only for very large $x$ ($x$ tending
towards infinity), we can create the following overview:
$$
\begin{array}{ccccc}
x     &  ln(x)  &  x / ln(x) & PI(x)(counted) &       PI(x) / (x/ln(x)) \\
10^3  &  6.908  &   144      &  168        &       1.160 \\
10^6  &  13.816 &   72,386    &  78,498          &       1.085 \\
10^9  & 20.723  &   48,254,942 &  50,847,534       &       1.054
\end{array}
$$

>>>>

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, Jan. 2010, Aug. 2013, Mar. 2016)

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 20+ largest known primes (as of January 2016)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{prime number!records}

The following table contains the biggest, currently known primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet
at   \url{http://primes.utm.edu/largest.html},
at   \url{http://primes.utm.edu/mersenne/index.html}, and
at   \url{http://www.mersenne.org/primes/}.
}:

@@

\begin{table} % don't use [h] (error if table has >24 lines)
\begin{center}
\begin{tabular}{|c|cccc|}
\hline \rule{0pt}{10pt}    % Eyecatcher_neue-Mersenne
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline \rule{0pt}{15pt}
	1  & $2^{74,207,281}-1$ & 22,338,618 & 2016 & Mersenne, 49th known \\
	2  & $2^{57,885,161}-1$ & 17,425,170 & 2013 & Mersenne, 48th known \\
	3  & $2^{43,112,609}-1$ & 12,978,189 & 2008 & Mersenne, 47th known \\
	4  & $2^{42,643,801}-1$ & 12,837,064 & 2009 & Mersenne, 46th known \\
	5  & $2^{37,156,667}-1$ & 11,185,272 & 2008 & Mersenne, 45th known \\
	6  & $2^{32,582,657}-1$ &  9,808,358 & 2006 & Mersenne, M-44 \\
	7  & $2^{30,402,457}-1$ &  9,152,052 & 2005 & Mersenne, M-43 \\
	8  & $2^{25,964,951}-1$ &  7,816,230 & 2005 & Mersenne, M-42 \\
	9  & $2^{24,036,583}-1$ &  7,235,733 & 2004 & Mersenne, M-41 \\
	10 & $2^{20,996,011}-1$ &  6,320,430 & 2003 & Mersenne, M-40 \\
	11 & $2^{13,466,917}-1$ &  4,053,946 & 2001 & Mersenne, M-39 \\
	12 & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\
	13 & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\
	14 & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\
	15  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
	16  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
	17  & $4,847 \cdot 2^{3,321,063}+1$ & 999,744 & 2005 & Generalized Mersenne \\
	18  & $3 \cdot 2^{3,136,255}-1$ & 944,108 & 2007 & Generalized Mersenne \\
	19  & $2^{ 3,021,377}-1$ &   909,526 & 1998 & Mersenne, M-37 \\
	20  & $2^{ 2,976,221}-1$ &   895,932 & 1997 & Mersenne, M-36 \\
	21  & $222,361 \cdot 2^{2,854,840}+1$ & 859,398 & 2006 & Generalized Mersenne \\
	22 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\
	23  & $1,361,244^{131,072}+1$ &   803,988 & 2004 & Generalized Fermat \\
	24  & $1,176,694^{131,072}+1$ &   795,695 & 2003 & Generalized Fermat \\
	25  & $342,673 \cdot 2^{2,639,439}-1$ & 794,556 & 2007 & Generalized Mersenne \\

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of January 2016)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\addtocounter{footnote}{-1}
\footnotetext{\index{number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\url{http://www.seventeenorbust.com})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its task sometime.

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
Landon Curt Noll\index{Noll, Landon Curt} lists all known Mersenne primes
including its date of discovery and its value as number and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html} \\
Also see:
      \url{http://www.utm.edu/}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 42,643,801, ~ 43,112,609.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$49$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{prime number!Mersenne}\index{Mersenne!prime number}.

@@

\vskip +10 pt
For the first 44  % be_2005_UPDATEN_if-new-mersenne-prime-appears  % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 49th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely.\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/status.htm}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{prime number!test}. }

% be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Right now (March 2016) all prime exponents smaller than $ 35,183,273 $ have been
tested and double-checked\footnote{%
See home page of the GIMPS project:
      \url{http://www.mersenne.org/report_milestones}.
                                  }:
So we can
be certain, that this is really the 41th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime). Here are some examples in more detail:

@@

Right now the GIMPS project has discovered 15
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
largest Mersenne primes so far, including the largest known prime number at all.

@@

   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\begin{table}[ht]
\begin{center}
\begin{tabular}{|cccc|}
\hline \rule{0pt}{10pt} 
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline \rule{0pt}{15pt}

	$2^{74,207,281}-1$ & 22,338,618 & January 7, 2016 & Curtis Cooper  \\
	$2^{57,885,161}-1$ & 17,425,170 & January 25, 2013 & Curtis Cooper  \\
	$2^{43,112,609}-1$ & 12,978,189 & August 23, 2008 & Edson Smith  \\
	$2^{42,643,801}-1$ & 12,837,064 & April 12, 2009 & Odd Magnar Strindmo \\
	$2^{37,156,667}-1$ & 11,185,272 & September 6, 2008 & Hans-Michael Elvenich \\

@@

\hline
\end{tabular}
   % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
\caption{The largest primes found by the GIMPS project (as of January 2016)}
\label{Primes_GIMPS-Primes_table-reference}
\end{center}
\end{table}

@@

The formulae for the prime number theorem only apply when n tends to infinity.
The formula of Gauss can be replaced by more precise formulae.
For $x \geq 67$:
$$ ln(x) - 1,5 < x / PI(x) < ln(x) - 0,5 $$
Given that we know $PI(x)  =  x / \ln x$ only for very large $x$ ($x$ tending
towards infinity), we can create the following overview:
$$
\begin{array}{ccccc}
x     &  ln(x)  &  x / ln(x) & PI(x)(counted) &       PI(x) / (x/ln(x)) \\
10^3  &  6.908  &   144      &  168        &       1.160 \\
10^6  &  13.816 &   72,386    &  78,498          &       1.085 \\
10^9  & 20.723  &   48,254,942 &  50,847,534       &       1.054
\end{array}
$$
